---
title: "ensemble work"
author: "Andrea"
date: "2023-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ensemble


Ideally, for each type of model (i.e. each workflow), we want the best model. We
can get those with:

```{r}
rank_results(leopards_models, rank_metric = "roc_auc", select_best = TRUE)

```
And plot them with:
```{r}
autoplot(leopards_models, select_best=TRUE)
```

We are now ready to finalize the workflow_set, choosing the metric that we want to use:
```{r}
leopards_ensemble <- finalize_worflow_set(leopards_models, metric = "roc_auc")

leopards_ensemble <- fit(leopards_ensemble, data=leopards_training)

leopards_testing_preds <-predict(leopards_ensemble, new_data = leopards_testing, type="prob")

# we want to add a last_fit option for ensemble objects
```




# Troubleshooting

Sometimes, when turing a workflow_set, some individual resamples might fail. This
will be generally visible when creating summaries, as the n of the resamples will
be lower. We can investigate the problematic workflow with:
```{r}
extract_workflow_set_result(leopards_models, id = "default_gam")
```

