---
title: "tidysdm overview"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{tidysdm overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# set options to limit threads used by imported libraries
options(cores=2)
options(mc.cores=2)
# xgboost uses data.table
data.table::setDTthreads(2)
```

# SDMs with `tidymodels`

Species Distribution Modelling relies on several algorithms, many of
which have a number of hyperparameters that require turning. The
`tidymodels` universe includes a number of packages specifically design
to fit, tune and validate models. The advantage of `tidymodels` is that
the models syntax and the results returned to the users are
standardised, thus providing a coherent interface to modelling. Given
the variety of models required for SDM, `tidymodels` is an ideal
framework. `tidysdm` provides a number of wrappers and specialised
functions to facilitate the fitting of SDM with `tidymodels`.

This article provides an overview of the how `tidysdm` facilitates
fitting SDMs. Further articles, detailing how to use the package for
palaeodata, fitting more complex models and how to troubleshoot models
can be found on the [`tidisdm`
website](https://evolecolgroup.github.io/tidysdm/). As `tidysdm` relies
on `tidymodels`, users are advised to familiarise themselves with the
introductory tutorials on the [`tidymodels`
website](https://www.tidymodels.org/start/).

When we load `tidysdm`, it automatically loads `tidymodels` and all
associated packages necessary to fit models:

```{r libraries}
library(tidysdm)
```

### Accessing the data for this vignette: how to use `rgbif`

We start by reading in a set of presences for a species of lizard that
inhabits the Iberian peninsula, *Lacerta schreiberi*. This data is taken
from GBIF Occurrence Download (6 July 2023)
<https://doi.org/10.15468/dl.srq3b3>. The dataset is already included in
the `tidysdm` package:

```{r load_presences}
 data(lacerta)
 lacerta
```

Alternatively, we can easily access and manipulate this dataset using
`rbgif`:

```{r download_presences, eval = FALSE}
# download presences
library(rgbif)
occ_download_get(key = "0068808-230530130749713", path = tempdir())
# read file
library(readr)
distrib <- read_delim(file.path(tempdir(), "0068808-230530130749713.zip"))
# keep the necessary columns and rename them
lacerta <- distrib %>% select(gbifID, decimalLatitude, decimalLongitude) %>%
  rename(ID = gbifID, latitude = decimalLatitude, longitude = decimalLongitude)
```


# Preparing your data

First, let us visualise our presences by plotting on a map. `tidysdm`
works with `sf` objects to represent locations, so we will cast our
coordinates into an `sf` object, and set its projections to standard
'lonlat' (`crs` = 4326).

```{r cast_to_sf}
library(sf)
lacerta <- st_as_sf(lacerta, coords = c("longitude", "latitude"))
st_crs(lacerta) <- 4326
```

It is usually advisable to plot the locations directly on the raster
that will be used to extract climatic variables, to see how the
locations fall within the discrete space of the raster. For this
vignette, we will use WorldClim as our source of climatic information.
We will access the WorldClim data via the library `pastclim`; even
though this library, as the name suggests, is mostly designed to handle
palaeoclimatic reconstructions, it also provides convenient functions to
access present day reconstructions and future projections. `pastclim`
has a handy function to get the land mask for the available datasets,
which we can use as background for our locations. We will cut the raster
to the Iberian peninsula, where our lizard lives. For this simply
illustration, we will not bother to project the raster, but an equal
area projection would be desirable...

```{r land_mask, eval=FALSE}
library(pastclim)
download_dataset(dataset = "WorldClim_2.1_10m")
land_mask <-
  get_land_mask(time_ce = 1985, dataset = "WorldClim_2.1_10m")

# Iberia peninsula extension
iberia_poly <-
  terra::vect(
    "POLYGON((-9.8 43.3,-7.8 44.1,-2.0 43.7,3.6 42.5,3.8 41.5,1.3 40.8,0.3 39.5,
     0.9 38.6,-0.4 37.5,-1.6 36.7,-2.3 36.3,-4.1 36.4,-4.5 36.4,-5.0 36.1,
    -5.6 36.0,-6.3 36.0,-7.1 36.9,-9.5 36.6,-9.4 38.0,-10.6 38.9,-9.5 40.8,
    -9.8 43.3))"
  )

crs(iberia_poly) <- "lonlat"
# crop the extent
land_mask <- crop(land_mask, iberia_poly)
# and mask to the polygon
land_mask <- mask(land_mask, iberia_poly)
```

```{r echo=FALSE, results='hide'}
library(pastclim)
set_data_path(on_CRAN = TRUE)
# Iberia peninsula extension
iberia_poly <- terra::vect("POLYGON((-9.8 43.3,-7.8 44.1,-2.0 43.7,3.6 42.5,3.8 41.5,1.3 40.8,0.3 39.5,0.9 38.6,-0.4 37.5,-1.6 36.7,-2.3 36.3,-4.1 36.4,-4.5 36.4,-5.0 36.1,-5.6 36.0,-6.3 36.0,-7.1 36.9,-9.5 36.6,-9.4 38.0,-10.6 38.9,-9.5 40.8,-9.8 43.3))")

crs(iberia_poly) <- "lonlat"
gdal(warn = 3)
land_mask <- rast(system.file("extdata/lacerta_land_mask.nc",
  package = "tidysdm"
))
```

For plotting, we will take advantage of `tidyterra`, which makes
handling of `terra` rasters with `ggplot` a breeze.

```{r, fig.width=6, fig.height=4}
library(tidyterra)
library(ggplot2)
ggplot() +
  geom_spatraster(data = land_mask, aes(fill = land_mask_1985)) +
  geom_sf(data = lacerta) + scale_fill_gradient(na.value = "transparent")

```


# Thinning step

Now, we thin the observations to have one per cell in the raster (it
would be better if we had an equal area projection...):

```{r thin_by_cell}
set.seed(1234567)
lacerta <- thin_by_cell(lacerta, raster = land_mask)
nrow(lacerta)
```

```{r plot_thin_by_cell, fig.width=6, fig.height=4}
ggplot() +
  geom_spatraster(data = land_mask, aes(fill = land_mask_1985)) +
  geom_sf(data = lacerta) + scale_fill_gradient(na.value = "transparent")
```

Now, we thin further to remove points that are closer than 20km.
However, note that the standard map units for a 'lonlat' projection are
meters. `tidysdm` provides a convening conversion function, `km2m()`, to
avoid having to write lots of zeroes):

```{r thin_by_dist}
set.seed(1234567)
lacerta_thin <- thin_by_dist(lacerta, dist_min = km2m(20))
nrow(lacerta_thin)
```

Let's see what we have left of our points:

```{r plot_thin_by_dist, fig.width=6, fig.height=4}
ggplot() +
  geom_spatraster(data = land_mask, aes(fill = land_mask_1985)) +
  geom_sf(data = lacerta_thin) + scale_fill_gradient(na.value = "transparent")
```

We now need to select points that represent the potential available area
for the species. There are two approaches, we can either sample the
background with `sample_background()`, or we can generate
pseudo-absences with `sample_pseudoabs()`. In this example, we will
sample the background; more specifically, we will attempt to account for
potential sampling biases by using a target group approach, where
presences from other species within the same taxonomic group are used to
condition the sampling of the background, providing information on
differential sampling of different areas within the region of interest.

We will start by downloading records from 8 genera of *Lacertidae*,
covering the same geographic region of the Iberian peninsula from GBIF
<https://doi.org/10.15468/dl.53js5z>:

```{r download_background, eval=FALSE}
# download presences
library(rgbif)
# download file
occ_download_get(key = "0121761-240321170329656", path = tempdir())
# read file
library(readr)
backg_distrib <- readr::read_delim(file.path(tempdir(), "0121761-240321170329656.zip"))

# keep the necessary columns
lacertidae_background <- backg_distrib %>% select(gbifID, decimalLatitude, decimalLongitude) %>%
  rename(ID = gbifID, latitude = decimalLatitude, longitude = decimalLongitude)

lacertidae_background <- st_as_sf(lacertidae_background, coords = c("longitude", "latitude"))
st_crs(lacertidae_background) <- 4326
```


```{r echo=FALSE}
data("lacertidae_background")
lacertidae_background <- st_as_sf(lacertidae_background, coords = c("longitude", "latitude"))
st_crs(lacertidae_background) <- 4326
```

We need to convert these observations into a raster whose values are the
number of records (which will be later used to determine how likely each
cell is to be used as a background point):

```{r background_to_raster}
lacertidae_background_raster <- rasterize(lacertidae_background, land_mask, fun = "count")

plot(lacertidae_background_raster)

```

We can see that the sampling is far from random, with certain locations
having very large number of records. We can now sample the background,
using the 'bias' method to represent this heterogeneity in sampling
effort:

```{r sample_background}
set.seed(1234567)
lacerta_thin <- sample_background(data = lacerta_thin, raster = lacertidae_background_raster,
                  n = 3 * nrow(lacerta_thin),
                  method = "bias",
                  class_label = "background",
                  return_pres = TRUE)

```

Let's see our presences and background:

```{r plot_sample_pseudoabs, fig.width=6, fig.height=4}
ggplot() +
  geom_spatraster(data = land_mask, aes(fill = land_mask_1985)) +
  geom_sf(data = lacerta_thin, aes(col = class)) + scale_fill_gradient(na.value = "transparent")
```

Generally, we can use `pastclim` to check what variables are available
for the WorldClim dataset:

```{r load_climate, eval=FALSE}
climate_vars <- get_vars_for_dataset("WorldClim_2.1_10m")
```

```{r echo=FALSE, results='hide'}
gdal(warn = 3)
climate_present <- rast(system.file("extdata/lacerta_climate_present_10m.nc",
  package = "tidysdm"
))
climate_vars <- names(climate_present)
```

We first download the dataset at the right resolution (here 10
arc-minutes):

```{r eval=FALSE}
download_dataset("WorldClim_2.1_10m")
```

And then create a `terra` `SpatRaster` object. The dataset covers the
period 1970-2000, so `pastclim` dates it as 1985 (the midpoint). We can
directly crop to the Iberian peninsula:

```{r climate_worldclim, eval=FALSE}
climate_present <- pastclim::region_slice(
  time_ce = 1985,
  bio_variables = climate_vars,
  data = "WorldClim_2.1_10m",
  crop = iberia_poly
)
```

Next, we extract climate for all presences and background points:

```{r}
lacerta_thin <- lacerta_thin %>%
  bind_cols(terra::extract(climate_present, lacerta_thin, ID = FALSE))
```

Based on this paper (<https://doi.org/10.1007/s10531-010-9865-2>), we
are interested in these variables: "bio06", "bio05", "bio13", "bio14", "bio15".
We can visualise the differences between presences and the background using violin plots:

```{r fig.height=11, fig.width=7}
lacerta_thin %>% plot_pres_vs_bg(class)
```
We can see that all the variables of interest do seem to have a different distribution
between presences and the background. We can formally quantify the mismatch between the two 
by computing the overlap:

```{r}
lacerta_thin %>% dist_pres_vs_bg(class)
```

Again, we can see that the variables of interest seem good candidates with a clear
signal. Let us then focus on those variables:

```{r climate_variables}

suggested_vars <- c("bio06", "bio05", "bio13", "bio14", "bio15")
```

Environmental variables are often highly correlated, and collinearity is
an issue for several types of models. We can inspect the correlation
among variables with:

```{r, fig.width=7, fig.height=8}
pairs(climate_present[[suggested_vars]])

```

We can see that some variables have rather high correlation (e.g. bio05
vs bio14). We can subset to variables below a certain threshold
correlation (e.g. 0.7) with:

```{r choose_var_cor_keep}
climate_present <- climate_present[[suggested_vars]]

vars_uncor <- filter_collinear(climate_present, cutoff = 0.7, method = "cor_caret")
vars_uncor
```

So, removing bio14 leaves us with a set of uncorrelated variables. Note that
`filter_collinear` has other methods based on variable inflation that would also
be worth exploring. For this examle, we will remove bio14 and work with the remaining
variables.

```{r}
lacerta_thin <- lacerta_thin %>% select(all_of(c(vars_uncor, "class")))
climate_present <- climate_present[[vars_uncor]]
names(climate_present) # added to highlight which variables are retained in the end
```

# Fit the model by cross-validation

Next, we need to set up a `recipe` to define how to handle our dataset.
We don't want to do anything to our data in terms of transformations, so
we just need to define the formula (*class* is the `outcome`, all other
variables are `predictors`; note that, for `sf` objects, `geometry` is
automatically replaced by `X` and `Y` columns which are assigned a role
of `coords`, and thus not used as predictors):

```{r recipe}
lacerta_rec <- recipe(lacerta_thin, formula = class ~ .)
lacerta_rec
```

In classification models for `tidymodels`, the assumption is that the
level of interest for the response (in our case, presences) is the
reference level. We can confirm that we have the data correctly
formatted with:

```{r}
lacerta_thin %>% check_sdm_presence(class)
```

We now build a `workflow_set` of different models, defining which
hyperparameters we want to tune. We will use *glm*, *random forest*,
*boosted_trees* and *maxent* as our models (for more details on how to
use `workflow_set`s, see [this
tutorial](https://workflowsets.tidymodels.org/articles/tuning-and-comparing-models.html)).
The latter three models have tunable hyperparameters. For the most
commonly used models, `tidysdm` automatically chooses the most important
parameters, but it is possible to fully customise model specifications
(e.g. see the help for `sdm_spec_rf`).

```{r workflow_set}
lacerta_models <-
  # create the workflow_set
  workflow_set(
    preproc = list(default = lacerta_rec),
    models = list(
      # the standard glm specs
      glm = sdm_spec_glm(),
      # rf specs with tuning
      rf = sdm_spec_rf(),
      # boosted tree model (gbm) specs with tuning
      gbm = sdm_spec_boost_tree(),
      # maxent specs with tuning
      maxent = sdm_spec_maxent()
    ),
    # make all combinations of preproc and models,
    cross = TRUE
  ) %>%
  # tweak controls to store information needed later to create the ensemble
  option_add(control = control_ensemble_grid())
```

We now want to set up a spatial block cross-validation scheme to tune
and assess our models. We will split the data by creating 3 folds. We
use the `spatial_block_cv` function from the package `spatialsample`.
`spatialsample` offers a number of sampling approaches for spatial data;
it is also possible to convert objects created with `blockCV` (which
offers further features for spatial sampling, such as stratified
sampling) into an `rsample` object suitable to `tisysdm` with the
function `blockcv2rsample`.

```{r training_cv, fig.width=6, fig.height=4}
library(tidysdm)
set.seed(100)
#lacerta_cv <- spatial_block_cv(lacerta_thin, v = 5)

lacerta_cv <- spatial_block_cv(data = lacerta_thin, v = 3, n = 5)
autoplot(lacerta_cv)
```

We can now use the block CV folds to tune and assess the models (to keep
computations fast, we will only explore 3 combination of hyperparameters
per model; this is far too little in real life!):

```{r tune_grid}
set.seed(1234567)
lacerta_models <-
  lacerta_models %>%
  workflow_map("tune_grid",
    resamples = lacerta_cv, grid = 3,
    metrics = sdm_metric_set(), verbose = TRUE
  )
```

Note that `workflow_set` correctly detects that we have no tuning
parameters for *glm*. We can have a look at the performance of our
models with:

```{r autoplot_models, fig.width=7, fig.height=4}
autoplot(lacerta_models)
```

Now let's create an ensemble, selecting the best set of parameters for
each model (this is really only relevant for the random forest, as there
were not hype-parameters to tune for the *glm* and *gam*). We will use
the Boyce continuous index as our metric to choose the best random
forest and boosted tree. When adding members to an ensemble, they are
automatically fitted to the full training dataset, and so ready to make
predictions.

```{r}
lacerta_ensemble <- simple_ensemble() %>%
  add_member(lacerta_models, metric = "boyce_cont")
lacerta_ensemble

```

And visualise it

```{r autoplot_ens, fig.width=7, fig.height=4}
autoplot(lacerta_ensemble)
```

A tabular form of the model metrics can be obtained with:

```{r}
lacerta_ensemble %>% collect_metrics()
```

# Projecting to the present

We can now make predictions with this ensemble (using the default option
of taking the mean of the predictions from each model).

```{r plot_present, fig.width=6, fig.height=4}
prediction_present <- predict_raster(lacerta_ensemble, climate_present)
ggplot() +
  geom_spatraster(data = prediction_present, aes(fill = mean)) +
  scale_fill_terrain_c() +
  # plot presences used in the model
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```

We can subset the ensemble to only use the best models, based on the
Boyce continuous index, by setting a minimum threshold of 0.7 for that
metric. We will also take the median of the available model predictions
(instead of the mean, which is the default). The plot does not change
much (the models are quite consistent).

```{r plot_present_best, fig.width=6, fig.height=4}

prediction_present_boyce <- predict_raster(lacerta_ensemble, climate_present,
  metric_thresh = c("boyce_cont", 0.7),
  fun = "median"
)
ggplot() +
  geom_spatraster(data = prediction_present_boyce, aes(fill = median)) +
  scale_fill_terrain_c() +
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```

Sometimes, it is desirable to have binary predictions (presence vs
absence), rather than the probability of occurrence. To do so, we first
need to calibrate the threshold used to convert probabilities into
classes (in this case, we optimise the TSS):


```{r}
lacerta_ensemble <- calib_class_thresh(lacerta_ensemble,
  class_thresh = "tss_max", 
  metric_thresh = c("boyce_cont", 0.7)
)
```

And now we can predict for the whole continent:

```{r, fig.width=6, fig.height=4}
prediction_present_binary <- predict_raster(lacerta_ensemble,
  climate_present,
  type = "class",
  class_thresh = c("tss_max"), 
  metric_thresh = c("boyce_cont", 0.7)
)
ggplot() +
  geom_spatraster(data = prediction_present_binary, aes(fill = binary_mean)) +
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```

# Projecting to the future

WorldClim has a wide selection of projections for the future based on
different models and Shared Socio-economic Pathways (SSP). Type
`help("WorldClim_2.1")` for a full list. We will use predictions based
on "HadGEM3-GC31-LL" model for SSP 245 (intermediate green house gas
emissions) at the same resolution as the present day data (10
arc-minutes). We first download the data:

```{r eval=FALSE}
download_dataset("WorldClim_2.1_HadGEM3-GC31-LL_ssp245_10m")
```

Let's see what times are available:

```{r eval=FALSE}
get_time_ce_steps("WorldClim_2.1_HadGEM3-GC31-LL_ssp245_10m")
```

```{r echo=FALSE}
c(2030, 2050, 2070, 2090)
```

We will predict for 2090, the further prediction in the future that is
available.

Let's now check the available variables:

```{r eval=FALSE}
get_vars_for_dataset("WorldClim_2.1_HadGEM3-GC31-LL_ssp245_10m")
```

```{r echo=FALSE}
climate_vars[-length(climate_vars)]
```

Note that future predictions do not include *altitude* (as that does not
change with time), so if we needed it, we would have to copy it over
from the present. However, it is not in our set of uncorrelated
variables that we used earlier, so we don't need to worry about it.

```{r eval=FALSE}
climate_future <- pastclim::region_slice(
  time_ce = 2090,
  bio_variables = vars_uncor,
  data = "WorldClim_2.1_HadGEM3-GC31-LL_ssp245_10m",
  crop = iberia_poly
)
```

```{r echo=FALSE, results='asis'}
gdal(warn = 3)
climate_future <- rast(system.file("extdata/lacerta_climate_future_10m.nc",
  package = "tidysdm"
))
```

And predict using the ensemble:

```{r plot_future, fig.width=6, fig.height=4}
prediction_future <- predict_raster(lacerta_ensemble, climate_future)

ggplot() +
  geom_spatraster(data = prediction_future, aes(fill = mean)) +
  scale_fill_terrain_c()
```

# Visualising the contribution of individual variables

It is sometimes of interest to understand the relative contribution of
individual variables to the prediction. This is a complex task,
especially if there are interactions among variables. For simpler linear
models, it is possible to obtain marginal response curves (which show
the effect of a variable whilst keeping all other variables to their
mean) using `step_profile()` from the `recipes` package. We use
`step_profile()` to define a new recipe which we can then bake to
generate the appropriate dataset to make the marginal prediction. We can
then plot the predictions against the values of the variable of
interest. For example, to investigate the contribution of `bio05`, we
would:

```{r}
bio05_prof <- lacerta_rec %>%
  step_profile(-bio05, profile = vars(bio05)) %>%
  prep(training = lacerta_thin)

bio05_data <- bake(bio05_prof, new_data = NULL)

bio05_data <- bio05_data %>%
  mutate(
    pred = predict(lacerta_ensemble, bio05_data)$mean
  )

ggplot(bio05_data, aes(x = bio05, y = pred)) +
  geom_point(alpha = .5, cex = 1)
```

It is also possible to use
[DALEX](https://modeloriented.github.io/DALEX/),to explore `tidysdm`
models; see more details in the [tidymodels
additions](https://evolecolgroup.github.io/tidysdm/dev/articles/a2_tidymodels_additions.html)
article.

# Repeated ensembles

The steps of thinning and sampling pseudo-absences can have a bit impact
on the performance of SDMs. As these steps are stochastic, it is good
practice to explore their effect by repeating them, and then creating
ensembles of models over these repeats. In `tidysdm`, it is possible to
create `repeat_ensembles`. We start by creating a list of
`simple_ensembles`, by looping through the SDM pipeline. We will just
use two fast models to speed up the process.

```{r}
# empty object to store the simple ensembles that we will create
ensemble_list <- list()
set.seed(123) # make sure you set the seed OUTSIDE the loop
for (i_repeat in 1:3) {
  # thin the data
  lacerta_thin_rep <- thin_by_cell(lacerta, raster = climate_present)
  lacerta_thin_rep <- thin_by_dist(lacerta_thin_rep, dist_min = 20000)
  # sample pseudo-absences
  lacerta_thin_rep <- sample_pseudoabs(lacerta_thin_rep,
    n = 3 * nrow(lacerta_thin_rep),
    raster = climate_present,
    method = c("dist_min", 50000)
  )
  # get climate
  lacerta_thin_rep <- lacerta_thin_rep %>%
    bind_cols(terra::extract(climate_present, lacerta_thin_rep, ID = FALSE))
  # create folds
  lacerta_thin_rep_cv <- spatial_block_cv(lacerta_thin_rep, v = 5)
  # create a recipe
  lacerta_thin_rep_rec <- recipe(lacerta_thin_rep, formula = class ~ .)
  # create a workflow_set
  lacerta_thin_rep_models <-
    # create the workflow_set
    workflow_set(
      preproc = list(default = lacerta_thin_rep_rec),
      models = list(
        # the standard glm specs
        glm = sdm_spec_glm(),
        # maxent specs with tuning
        maxent = sdm_spec_maxent()
      ),
      # make all combinations of preproc and models,
      cross = TRUE
    ) %>%
    # tweak controls to store information needed later to create the ensemble
    option_add(control = control_ensemble_grid())

  # train the model
  lacerta_thin_rep_models <-
    lacerta_thin_rep_models %>%
    workflow_map("tune_grid",
      resamples = lacerta_thin_rep_cv, grid = 10,
      metrics = sdm_metric_set(), verbose = TRUE
    )
  # make an simple ensemble and add it to the list
  ensemble_list[[i_repeat]] <- simple_ensemble() %>%
    add_member(lacerta_thin_rep_models, metric = "boyce_cont")
}
```

Now we can create a `repeat_ensemble` from the list:

```{r}
lacerta_rep_ens <- repeat_ensemble() %>% add_repeat(ensemble_list)
lacerta_rep_ens
```

We can summarise the goodness of fit of models for each repeat with
`collect_metrics()`, but there is no `autoplot()` function for
`repeated_ensemble` objects.

We can then predict in the usual way (we will take the mean and median
of all models):

```{r, fig.width=6, fig.height=4}
lacerta_rep_ens <- predict_raster(lacerta_rep_ens, climate_present,
  fun = c("mean", "median")
)
ggplot() +
  geom_spatraster(data = lacerta_rep_ens, aes(fill = median)) +
  scale_fill_terrain_c()
```

# Multivariate environmental similarity surfaces (MESS)

The total area of projection of the model may include environmental conditions 
which lie outside the range of conditions covered by the calibration dataset. 
This phenomenon can lead to misinterpretation of the SDM outcomes due to 
spatial extrapolation. 

Multivariate environmental similarity surfaces (MESS) (Elith et al. 2010) can be
used to highlight areas where extrapolation occurs and thus provide details 
regarding the prediction's uncertainty.

Firstly, we can estimate MESS for present predictions

```{r}
lacerta_mess_present <- extrapol_mess(x = climate_present, training = lacerta_thin, .col = "class")

ggplot() + geom_spatraster(data = lacerta_mess_present) + scale_fill_viridis_b(na.value = "transparent")
```

Extrapolation occurs in areas where MESS values are negative. 

We can now overlay MESS values with current prediction to visualize areas characterized by spatial extrapolation. 

```{r}
# subset mess 
lacerta_mess_present_subset <- lacerta_mess_present
lacerta_mess_present_subset[lacerta_mess_present_subset >= 0] <- NA
lacerta_mess_present_subset[lacerta_mess_present_subset < 0] <- 1

# convert into polygon
lacerta_mess_present_subset <- as.polygons(lacerta_mess_present_subset)

# plot as a mask 
terra::plot(prediction_present, fun = function() polys(lacerta_mess_present_subset, col = "black", lwd = 0.5))
```

Estimating MESS it is also useful for assess for extrapolation among different time periods, where novel climatic conditions may arise. 

We can compute MESS for future predictions in the same way we have done previously

```{r}
lacerta_mess_future <- extrapol_mess(x = climate_future, training = lacerta_thin, .col = "class")

ggplot() + geom_spatraster(data = lacerta_mess_future) + scale_fill_viridis_b(na.value = "transparent")
```

We can now mask areas of future prediction characterised by negative MESS values 

```{r}
# subset mess 
lacerta_mess_future_subset <- lacerta_mess_future
lacerta_mess_future_subset[lacerta_mess_future_subset >= 0] <- NA
lacerta_mess_future_subset[lacerta_mess_future_subset < 0] <- 1

# convert into polygon
lacerta_mess_future_subset <- as.polygons(lacerta_mess_future_subset)

# plot as a mask 
terra::plot(prediction_future, fun = function() polys(lacerta_mess_future_subset, col = "black", lwd = 0.5))
```

@TODO maybe need to make the mask more transparent?

It is is important to keep in mind that MESS provide a scale of values, which means that not all the areas characterised by negative MESS values have the same level of extrapolation. 

@TODO maybe a brief mention within this section on the use of area of applicability?

