---
title: "Examples of additional tidymodels features"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2

vignette: >
  %\VignetteIndexEntry{Examples of additional tidymodels features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# xgboost uses data.table
data.table::setDTthreads(2)
RhpcBLASctl::blas_set_num_threads(2)
RhpcBLASctl::omp_set_num_threads(2)

download_data <- FALSE
# note that data were created in the overview vignette
```


In this vignette, we illustrate how a number of features from `tidymodels` can
be used to enhance a conventional SDM pipeline. If you’re new to `tidymodels`;
there are a number of excellent tutorials (both introductory and advanced) on
its dedicated [website](https://www.tidymodels.org/). We reuse the example on
the Iberian lizard from the [`tidysdm`
overview](https://evolecolgroup.github.io/tidysdm/articles/a0_tidysdm_overview.html)
article. Each subsection is written to be self-contained, so you can read and
run them independently.

# Using multi-level factors as predictors

Most machine learning algorithms do not natively use multilevel factors as
predictors. The solution is to create dummy variables, which are binary
variables that represent the levels of the factor. In `tidymodels`, this is
done using the `step_dummy` function.

Let's create a factor variable with 3 levels based on altitude.
```{r}
library(tidysdm)
library(tidyterra)
# load the dataset
lacerta_thin <- readRDS(system.file("extdata/lacerta_thin_all_vars.rds",
  package = "tidysdm"
))
# create a topography variable with 3 levels based on altitude
lacerta_thin$topography <- cut(lacerta_thin$altitude,
  breaks = c(-Inf, 200, 800, Inf),
  labels = c("plains", "hills", "mountains")
)
table(lacerta_thin$topography)
```

We then create the recipe by adding a step to create dummy variables for the
`topography` variable.
```{r}
# subset to variable of interest
lacerta_thin <- lacerta_thin %>% select(
  class, bio05, bio06, bio12,
  bio15, topography
)

lacerta_rec <- recipe(lacerta_thin, formula = class ~ .) %>%
  step_dummy(topography)
lacerta_rec
```

Let's us see what this does:
```{r}
lacerta_prep <- prep(lacerta_rec)
summary(lacerta_prep)
```
We have added two "derived" variables, *topography_hills* and
*topography_mountains*, which are binary variables that allow us to code
topography (with plains being used as the reference level, which is coded by
both hills and mountains being 0 for a given location). We can look at the
first few rows of the data to see the new variables by baking the recipe:

```{r}
lacerta_bake <- bake(lacerta_prep, new_data = lacerta_thin)
glimpse(lacerta_bake)
```

We can now run the sdm as usual:
```{r}
# define the models
lacerta_models <-
  # create the workflow_set
  workflow_set(
    preproc = list(default = lacerta_rec),
    models = list(
      # the standard glm specs
      glm = sdm_spec_glm(),
      # rf specs with tuning
      rf = sdm_spec_rf()
    ),
    # make all combinations of preproc and models,
    cross = TRUE
  ) %>%
  # tweak controls to store information needed later to create the ensemble
  option_add(control = control_ensemble_grid())
# tune
set.seed(100)
lacerta_cv <- spatial_block_cv(lacerta_thin, v = 3)
lacerta_models <-
  lacerta_models %>%
  workflow_map("tune_grid",
    resamples = lacerta_cv, grid = 3,
    metrics = sdm_metric_set(), verbose = TRUE
  )
# fit the ensemble
lacerta_ensemble <- simple_ensemble() %>%
  add_member(lacerta_models, metric = "boyce_cont")
```

We can now verify that the dummy variables were used by extracting the model
fit from one of the models in the ensemble:
```{r}
lacerta_ensemble$workflow[[1]] %>% extract_fit_parsnip()
```

We can see that we have coefficients for *topography_hills* and
*topography_mountains*.

Let us now predict the presence of the lizard in the Iberian Peninsula using
the ensemble. Note that, for `predict_raster()` to work, the name and levels
for a categorical variable need to match with those used when training the
models (i.e. in the recipe with `step_dummy()`):

```{r add_levels, fig.width=7, fig.height=4}
climate_present <- terra::readRDS(
  system.file("extdata/lacerta_climate_present_10m.rds",
    package = "tidysdm"
  )
)

# project climate layers
iberia_proj4 <-
  "+proj=aea +lon_0=-4.0 +lat_1=36.8 +lat_2=42.6 +lat_0=39.7 +datum=WGS84 +units=m +no_defs"
climate_present <- terra::project(climate_present, y = iberia_proj4)

# first we add a topography variable to the climate data
climate_present$topography <- climate_present$altitude
climate_present$topography <- terra::classify(climate_present$topography,
  rcl = c(-Inf, 200, 800, Inf),
  include.lowest = TRUE,
  brackets = TRUE
)
library(terra)
levels(climate_present$topography) <-
  data.frame(ID = c(0, 1, 2), topography = c("plains", "hills", "mountains"))

# now we can predict
predict_factor <- predict_raster(lacerta_ensemble, climate_present)

# and plot
ggplot() +
  geom_spatraster(data = predict_factor, aes(fill = mean)) +
  scale_fill_terrain_c() +
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```

```{r cleanup_multilevel_factors, include=FALSE}
# Clean up objects created in the "Using multi-level factors as predictors"
# section
rm(
  lacerta_thin,
  lacerta_rec,
  lacerta_prep,
  lacerta_bake,
  lacerta_models,
  lacerta_cv,
  lacerta_ensemble,
  climate_present,
  predict_factor
)
```

# Exploring models with `DALEX`

An issue with machine learning algorithms is that it is not easy to understand
the role of different variables in giving the final prediction. A number of
packages have been created to explore and explain the behaviour of ML
algorithms, such as those used in `tidysdm`. In the [`tidysdm`
overview](https://evolecolgroup.github.io/tidysdm/articles/a0_tidysdm_overview.html)
article, we illustrated how to use `recipes` to create profiles.

Here we demonstrate how to use [DALEX](https://modeloriented.github.io/DALEX/),
an excellent package that has methods to deal with `tidymodels`. `tidysdm`
contains additional functions that allow us to use the DALEX functions directly
on `tidysdm` ensembles.

We will use a simple ensemble that we built in the overview vignette.

```{r}
library(tidysdm)
lacerta_ensemble
```

The first step in DALEX is to create an explainer object, which can then be
queried by different functions in the package, to turn the explainer into an
explanation (following the DALEX lingo). As a first step, we use the custom
function `explain_tidysdm` to generate our explainer:
```{r}
explainer_lacerta_ens <- explain_tidysdm(lacerta_ensemble)
```
Now that we have our explainer, we can explore variable importance for the
ensemble:

```{r vip, fig.width=7, fig.height=4}
library(DALEX)
vip_ensemble <- model_parts(explainer = explainer_lacerta_ens)
plot(vip_ensemble)
```

Or generate partial dependency plots for a given variable (e.g. bio05):
```{r pdp, fig.width=7, fig.height=4}
pdp_bio05 <- model_profile(explainer_lacerta_ens, N = 500, variables = "bio05")
plot(pdp_bio05)
```

There are many other functions in DALEX that can be applied to the explainer to
further explore the behaviour of the model; see several tutorial on
https://modeloriented.github.io/DALEX/ for more details.

It is also possible to explore the individual models that make up the ensemble:
```{r}
explainer_list <- explain_tidysdm(tidysdm::lacerta_ensemble, by_workflow = TRUE)
```

The resulting list can be then used to build lists of explanations, which can
then be plotted.

```{r profile, fig.width=7, fig.height=4}
profile_list <- lapply(explainer_list, model_profile,
  N = 500,
  variables = "bio05"
)
plot(profile_list)
```

```{r cleanup_dalex, include=FALSE}
# Clean up objects created in the "Exploring models with DALEX" section
rm(
  explainer_lacerta_ens,
  vip_ensemble,
  pdp_bio05,
  explainer_list,
  profile_list
)
```

# Different recipes for certain models

Some model types (e.g. GLMs and SVMs) can be sensitive to multicollinearity
among predictors, while others (notably random forests) are generally more
robust to correlated inputs. In this section, we create two separate recipes:
one that keeps all predictors, and another that retains only a reduced set of
weakly correlated variables.

Load the data:

```{r}
library(tidysdm)
library(sf)
lacerta_thin <- readRDS(system.file("extdata/lacerta_thin_all_vars.rds",
  package = "tidysdm"
))
```

We then create two recipes: one that keeps all variables, and another that
removes highly correlated ones.

```{r two_recipes}
lacerta_rec_all <- recipe(lacerta_thin, formula = class ~ .)
lacerta_rec_uncor <- lacerta_rec_all %>%
  step_rm(all_of(c(
    "bio01", "bio02", "bio03", "bio04", "bio07", "bio08",
    "bio09", "bio10", "bio11", "bio12", "bio14", "bio16",
    "bio17", "bio18", "bio19", "altitude"
  )))

lacerta_rec_uncor
```

And now use these two recipes in a `workflowset` (we will keep it small for
computational time), selecting the appropriate recipe for each model. We will
include a model (polynomial support vector machines, or SVM) which does not
have a wrapper in `tidysdm` for creating a model specification. However, we can
use a standard model spec from `yardstick`:

```{r workflow_set_different_recipes}
lacerta_models <-
  # create the workflow_set
  workflow_set(
    preproc = list(
      uncor = lacerta_rec_uncor, # recipe for the glm
      all = lacerta_rec_all, # recipe for the random forest
      all = lacerta_rec_uncor # recipe for svm
    ),
    models = list(
      # the standard glm specs
      glm = sdm_spec_glm(),
      # rf specs with tuning
      rf = sdm_spec_rf(),
      # svm specs with tuning
      svm = parsnip::svm_poly(
        cost = tune::tune(),
        degree = tune::tune()
      ) %>%
        parsnip::set_engine("kernlab") %>%
        parsnip::set_mode("classification")
    ),
    # make all combinations of preproc and models,
    cross = FALSE
  ) %>%
  # tweak controls to store information needed later to create the ensemble
  # note that we use the bayes version as we will use a Bayes search (see later)
  option_add(control = stacks::control_stack_bayes())
```


We can now use the block CV folds to tune and assess the models. Note that
there are multiple tuning approaches, besides the standard grid method. Here we
will use `tune_bayes` from the `tune` package (see its help page to see how a
Gaussian Process model is used to choose parameter combinations).


This tuning method (as opposed to use a standard grid) does not allow for
hyper-parameters with unknown limits, but `mtry` for random forest is undefined
as its upper range depends on the number of variables in the dataset. So,
before tuning, we need to finalise `mtry` by informing the set dials with the
actual data:

```{r finalize_mtry}
set.seed(100)
# create the CV folds
lacerta_cv <- spatial_block_cv(lacerta_thin, v = 3)

# extract and finalize the rf parameters
rf_param <- lacerta_models %>%
  # extract the rf workflow
  extract_workflow("all_rf") %>%
  # extract its parameters dials (used to tune)
  extract_parameter_set_dials() %>%
  # give it the predictors to finalize mtry
  finalize(x = st_drop_geometry(lacerta_thin) %>% select(-class))

# now update the workflowset with the new parameter info
lacerta_models <- lacerta_models %>%
  option_add(param_info = rf_param, id = "all_rf")
```


And now we can tune the models:
```{r tune_grid_different_recipes}
set.seed(1234567)
lacerta_models <-
  lacerta_models %>%
  workflow_map("tune_bayes",
    resamples = lacerta_cv, initial = 8,
    metrics = sdm_metric_set(), verbose = TRUE
  )
```

We can have a look at the performance of our models with:

```{r autoplot_models, fig.width=7, fig.height=4}
autoplot(lacerta_models)
```

```{r cleanup_recipes_section, include=FALSE}
objs <- c(
  "lacerta_rec_all",
  "lacerta_rec_uncor",
  "lacerta_models",
  "rf_param",
  "lacerta_cv",
  "lacerta_thin"
)

rm(list = intersect(objs, ls()))
```

# The initial split {#initial-split}

The standard `tidymodels` workflow begins by splitting the data into a
**training set** (used to fit and tune models) and an **independent testing
set** (held out for final evaluation). This step is essential for getting an
honest estimate of predictive performance of any SDM, yet it has often been
omitted in traditional SDM practice, where model assessment has instead relied
primarily on (spatial) cross-validation. This is largely for historical
reasons: SDM datasets were frequently small, and holding data back for testing
felt too costly.

In this section, we demonstrate a robust approach by holding out 20% of the
data (1/5) for testing and using the remaining 80% for training. While we
strongly recommend this approach whenever the data allow, we acknowledge that
it can be data-hungry and may not always be feasible for smaller datasets. We
therefore recommend exploring multiple random seeds for the initial split to
assess the sensitivity of results to a particular partition, and to avoid
splitting the data when results are highly sensitive and unstable.

We start by loading a set of presences and absences and their associated
climate, analogous to the one that we generated in the [`tidysdm`
overview](https://evolecolgroup.github.io/tidysdm/articles/a0_tidysdm_overview.html)
article:

```{r load_data}
library(tidysdm)
library(sf)
library(tidyterra)
lacerta_thin <- readRDS(system.file("extdata/lacerta_thin_all_vars.rds",
  package = "tidysdm"
))
```

We then use `spatial_initial_split` to do the split, using a `spatial_block_cv`
scheme to partition the data:

```{r initial_split, fig.width=7, fig.height=4}
set.seed(1005)
lacerta_initial <- spatial_initial_split(lacerta_thin,
  prop = 1 / 5, spatial_block_cv
)
autoplot(lacerta_initial)
```

And check the balance of presences vs pseudoabsences:
```{r check_balance}
check_splits_balance(lacerta_initial, class)
```

Next, we need to set up a `recipe` to define how to handle our dataset.

```{r recipe_initial_split}
lacerta_rec <- recipe(lacerta_thin, formula = class ~ .)
lacerta_rec
```

And confirm that we have the data correctly formatted with:

```{r}
lacerta_thin %>% check_sdm_presence(class)
```

We can now extract the training set from our `lacerta_initial` split, and
sample folds to set up cross validation (note that we set the `cellsize` and
`offset` based on the full dataset, `lacerta_thin`; this allows us to use the
same grid we used for the `initial_split`). In this example, we also have to
add a very small number to the offset to avoid an error in the
`spatial_block_cv` function arising from some points falling on the boundary of
the grid cells). This is often not necessary, so only introduce this if you
encounter an error.

```{r training_cv, fig.width=7, fig.height=4}
set.seed(1005)
lacerta_training <- training(lacerta_initial)
lacerta_cv <- spatial_block_cv(lacerta_training,
  v = 5,
  cellsize = grid_cellsize(lacerta_thin),
  offset = grid_offset(lacerta_thin) + 0.00001
)
autoplot(lacerta_cv)
```

And check the balance in the dataset:
```{r}
check_splits_balance(lacerta_cv, class)
```

Same as in the [`tidysdm`
overview](https://evolecolgroup.github.io/tidysdm/articles/a0_tidysdm_overview.html)
article we can now build a `workflow_set` of different models:

```{r workflow_set_initial_split}
lacerta_models <-
  # create the workflow_set
  workflow_set(
    preproc = list(default = lacerta_rec),
    models = list(
      # the standard glm specs
      glm = sdm_spec_glm(),
      # rf specs with tuning
      rf = sdm_spec_rf(),
      # boosted tree model (gbm) specs with tuning
      gbm = sdm_spec_boost_tree(),
      # maxent specs with tuning
      maxent = sdm_spec_maxent()
    ),
    # make all combinations of preproc and models,
    cross = TRUE
  ) %>%
  # tweak controls to store information needed later to create the ensemble
  option_add(control = control_ensemble_grid())
```

And use the block CV folds to tune and assess the models:

```{r tune_grid_initial_split, fig.width=7, fig.height=4}
set.seed(1234567)
lacerta_models <-
  lacerta_models %>%
  workflow_map("tune_grid",
    resamples = lacerta_cv, grid = 3,
    metrics = sdm_metric_set(), verbose = TRUE
  )

autoplot(lacerta_models)
```

And create an ensemble, selecting the best set of parameters for each model
using Boyce continuous index as our metric:

```{r build_ensemble, fig.width=7, fig.height=4}
lacerta_ensemble <- simple_ensemble() %>%
  add_member(lacerta_models, metric = "boyce_cont")

autoplot(lacerta_ensemble)
```

We can now use the ensemble to make predictions about the testing data:

```{r predict_test_ensemble}
lacerta_testing <- testing(lacerta_initial)

lacerta_test_pred <-
  lacerta_testing %>%
  bind_cols(predict(lacerta_ensemble, ., type = "prob"))
```

And look at the goodness of fit:

```{r assess_test_ensemble}
sdm_metric_set()(data = lacerta_test_pred, truth = class, mean)
```

We can now make predictions with this ensemble using the best models.

```{r plot_present_best, fig.width=7, fig.height=4}
# load climate data
climate_present <- terra::readRDS(
  system.file("extdata/lacerta_climate_present_10m.rds",
    package = "tidysdm"
  )
)

# project climate layers
iberia_proj4 <-
  "+proj=aea +lon_0=-4.0 +lat_1=36.8 +lat_2=42.6 +lat_0=39.7 +datum=WGS84 +units=m +no_defs"
climate_present <- terra::project(climate_present, y = iberia_proj4)

# predict using only the best models
prediction_present_boyce <- predict_raster(lacerta_ensemble, climate_present,
  metric_thresh = c("boyce_cont", 0.5),
  fun = "median"
)

# plot
ggplot() +
  geom_spatraster(data = prediction_present_boyce, aes(fill = median)) +
  scale_fill_terrain_c() +
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```

# Stack ensembles

Now that we have an explicit training/testing split (see the [previous
section](#initial-split) section), we can go beyond a simple ensemble that
keeps only the single best configuration of each model type and averages their
predictions. Instead, we can build a stack ensemble using the `stacks` package.
Stacking fits a meta-learner that learns how to optimally weight predictions
from multiple candidate models, including multiple tuned versions of the same
algorithm with different hyper-parameters, so that the final ensemble can
exploit complementary strengths across model configurations.

An explicit testing split is not strictly required for stacking, since model
performance can be estimated using cross-validation. However, an independent
test set is recommended whenever an unbiased estimate of the final model’s
predictive performance is required. This is particularly important for stacked
ensembles, which can be prone to overfitting.

```{r build_stack, fig.width=7, fig.height=4}
library(stacks)
set.seed(1005)
lacerta_stack <-
  # initialize the stack
  stacks() %>%
  # add candidate members
  add_candidates(lacerta_models) %>%
  # determine how to combine their predictions
  blend_predictions(
    metric = sdm_metric_set() # use our SDM metrics for calculating stacking
    # coefficients
  ) %>%
  # fit candidates with non-zero weights (i.e. non-zero stacking coefficients)
  fit_members()

autoplot(lacerta_stack, type = "weights")
```

In this example, one random forest variant and one SVM variant were selected
(though more can be, depending on tuning and the data). The stacking
coefficients indicate how much weight each selected model contributes to the
final ensemble prediction. We can now use the ensemble to make predictions
about the testing data:

```{r predict_test_stack}
lacerta_testing <- testing(lacerta_initial)

lacerta_test_pred <-
  lacerta_testing %>%
  bind_cols(predict(lacerta_stack, ., type = "prob"))
```

And look at the goodness of fit using some commonly used sdm metrics. Note that
`sdm_metric_set` is first invoked to generate a function (with empty `()`) that
is then used on the data.

```{r assess_test_stack}
sdm_metric_set()(data = lacerta_test_pred, truth = class, .pred_presence)
```

We can now make predictions with this stacked ensemble. We start by extracting
the climate for the variables of interest:

```{r eval=download_data}
climate_present <- terra::readRDS(
  system.file("extdata/lacerta_climate_present_10m.rds",
    package = "tidysdm"
  )
)

# project climate layers
iberia_proj4 <-
  "+proj=aea +lon_0=-4.0 +lat_1=36.8 +lat_2=42.6 +lat_0=39.7 +datum=WGS84 +units=m +no_defs"
climate_present <- terra::project(climate_present, y = iberia_proj4)
```

```{r echo=FALSE, results='hide', eval=!download_data}
climate_vars <- lacerta_rec$var_info %>%
  filter(role == "predictor") %>%
  pull(variable)
if (!all(climate_vars %in% names(climate_present))) {
  stop("mismatched variables in the raster")
}
```
And predict the presence of the lizard using the stacked ensemble:

```{r plot_present, fig.width=7, fig.height=4}
prediction_present <- predict_raster(lacerta_stack, climate_present,
  type = "prob"
)
library(tidyterra)
ggplot() +
  geom_spatraster(data = prediction_present, aes(fill = .pred_presence)) +
  scale_fill_terrain_c() +
  # plot presences used in the model
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```
