---
title: "tidysdm workflow functionalities"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{tidysdm overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# set options to limit threads used by imported libraries
options(cores=2)
options(mc.cores=2)
# xgboost uses data.table
data.table::setDTthreads(2)
```

# SDMs with `tidymodels` : an overview of the workflow and functionality

[Intro/Abstract as in paper]

## Background

[intro to tidymodels]

## Workflow with `tidysdm`

### Accessing the data for this vignette: how to use `rgbif`

In this vignette, we use a set of presences for a species of lizard that inhabits the Iberian peninsula, *Lacerta schreiberi*. This data is taken from GBIF: . We can easily access and manipulate this data using `rbgif`. We illustrate here how to do it:

```{r download_presences}

# download presences
library(rgbif)
# download file
occ_download_get(key = "0068808-230530130749713", path = tempdir())

# unzip file
unzip(file.path(tempdir(), "0068808-230530130749713.zip"),  overwrite = TRUE,  exdir = tempdir(), unzip = "internal")

@TODO https://stackoverflow.com/questions/12460938/r-reading-in-a-zip-data-file-without-unzipping-it

# read file
library(readr)
# distrib <- data.table::fread(file.path(tempdir(),"0068808-230530130749713.csv"))
distrib <- read.table(file.path(tempdir(), "0068808-230530130749713.csv"),  row.names = NULL)
distrib <- readr::read_delim(file.path(tempdir(), "0068808-230530130749713.csv"))

# keep the necessary columns
lacerta <- distrib[, c("gbifID", "decimalLatitude", "decimalLongitude")]
names(lacerta) <- c("ID", "latitude", "longitude")


```

\

### Assembling data - before modelling step

explaining all the pre processing of data;

introduce the problem that we work with only presences and (sometimes) biased presences);

show different options of sampling background, pseudo-absences.

Download presences

Download climate

Thin

Sample background

\- sample background

\- sample pseudoabsences

Choice of variables

\- violinplot

Remove collinear variables

\- correlation

\- VIF

\- correlation and VIF

### Pre-processing of models / Modelling

[rationale for tidymodels: why are we using tidymodels, flexibility of having each step of the analysis under control (and with the option to parallelise for improving computing time) and easily add/chose various methods. ]

\
Creating workflow

\- recipe

\- model specifications

Create data folds

Tune the model

\- algorithms available

\- how to add additional algorithms (parsnip)

### **Evaluation**

Metrics available

### **Ensemble**

Simple ensemble

Stacked ensemble

Repeated ensemble

Ensemble from individual workflows

Averaging ensembles

Evaluation of ensembles

### **Projection**

Predict

Predict_raster 

\- same time slice

\- different time slice

Convert into binary

Compare niche

Managing extrapolation:

\- MESS

\- area of applicability (<https://rdrr.io/cran/waywiser/man/ww_area_of_applicability.html>)

\- clamping
