---
title: "Testing predict function for repeated ensembles"
output: html_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Libraries

```{r message=FALSE, warning=FALSE}
library(tidysdm)
library(sf)
library(patchwork)
library(stacks)
library(ggplot2)
library(tidyterra)
library(pastclim)
library(DALEX)
```

# Preparing data

We will use the same lacerta example as in the `tidysdm` vignettes.

```{r load_presences, include = FALSE}
data(lacerta)
lacerta <- st_as_sf(lacerta, coords = c("longitude", "latitude"))
st_crs(lacerta) <- "+proj=longlat"
```

```{r land_mask, include = FALSE}
download_dataset(dataset = "WorldClim_2.1_10m")
land_mask <-
  get_land_mask(time_ce = 1985, dataset = "WorldClim_2.1_10m")

# Iberia peninsula extension
iberia_poly <-
  terra::vect(
    "POLYGON((-9.8 43.3,-7.8 44.1,-2.0 43.7,3.6 42.5,3.8 41.5,1.3 40.8,0.3 39.5,
     0.9 38.6,-0.4 37.5,-1.6 36.7,-2.3 36.3,-4.1 36.4,-4.5 36.4,-5.0 36.1,
    -5.6 36.0,-6.3 36.0,-7.1 36.9,-9.5 36.6,-9.4 38.0,-10.6 38.9,-9.5 40.8,
    -9.8 43.3))"
  )

crs(iberia_poly) <- "+proj=longlat"
# crop the extent
land_mask <- crop(land_mask, iberia_poly)
# and mask to the polygon
land_mask <- mask(land_mask, iberia_poly)

iberia_proj4 <-
  "+proj=aea +lon_0=-4.0 +lat_1=36.8 +lat_2=42.6 +lat_0=39.7 +datum=WGS84 +units=m +no_defs"

land_mask <- terra::project(land_mask, y = iberia_proj4)
lacerta <- st_transform(lacerta, iberia_proj4)
```

```{r climate data, include = FALSE}
download_dataset("WorldClim_2.1_10m")
climate_vars <- get_vars_for_dataset("WorldClim_2.1_10m")
climate_present <- pastclim::region_slice(
  time_ce = 1985,
  bio_variables = climate_vars,
  data = "WorldClim_2.1_10m",
  crop = iberia_poly
)
climate_present <- terra::project(climate_present, y = iberia_proj4)

climate_vars <- names(climate_present)
```


# Classic repeated ensemble

```{r}
suggested_vars <- c("bio06", "bio05", "bio13", "bio14", "bio15")
climate_present_lac <- climate_present[[suggested_vars]]
# empty object to store the simple ensembles that we will create
ensemble_list <- list()
set.seed(1234) # make sure you set the seed OUTSIDE the loop
for (i_repeat in 1:3) {
  # select variables
  suggested_vars <- c("bio06", "bio05", "bio13", "bio14", "bio15")
  climate_present_lac <- climate_present[[suggested_vars]]
  # thin the data
  lacerta_thin_rep <- thin_by_cell(lacerta, raster = climate_present_lac)
  lacerta_thin_rep <- thin_by_dist(lacerta_thin_rep, dist_min = 20000)
  # sample pseudo-absences
  lacerta_thin_rep <- sample_pseudoabs(lacerta_thin_rep,
    n = 3 * nrow(lacerta_thin_rep),
    raster = climate_present_lac,
    method = c("dist_min", 50000)
  )
  # get climate
  lacerta_thin_rep <- lacerta_thin_rep %>%
    bind_cols(terra::extract(climate_present_lac, lacerta_thin_rep, ID = FALSE))
  # create folds
  lacerta_thin_rep_cv <- spatial_block_cv(lacerta_thin_rep, v = 5)
  # create a recipe
  lacerta_thin_rep_rec <- recipe(lacerta_thin_rep, formula = class ~ .)
  # create a workflow_set
  lacerta_thin_rep_models <-
    # create the workflow_set
    workflow_set(
      preproc = list(default = lacerta_thin_rep_rec),
      models = list(
        # the standard glm specs
        glm = sdm_spec_glm(),
        # maxent specs with tuning
        maxent = sdm_spec_maxent()
      ),
      # make all combinations of preproc and models,
      cross = TRUE
    ) %>%
    # tweak controls to store information needed later to create the ensemble
    option_add(control = control_ensemble_grid())

  # train the model
  lacerta_thin_rep_models <-
    lacerta_thin_rep_models %>%
    workflow_map("tune_grid",
      resamples = lacerta_thin_rep_cv, grid = 3,
      metrics = sdm_metric_set(), verbose = TRUE
    )
  # make an simple ensemble and add it to the list
  ensemble_list[[i_repeat]] <- simple_ensemble() %>%
    add_member(lacerta_thin_rep_models, metric = "boyce_cont")
}
```

Now we can create a `repeat_ensemble` from the list:

```{r}
lacerta_rep_ens <- repeat_ensemble() %>% add_repeat(ensemble_list)
lacerta_rep_ens
```

We can summarise the goodness of fit of models for each repeat with
`collect_metrics()`, but there is no `autoplot()` function for
`repeated_ensemble` objects.

We can then predict in the usual way. We will take the mean and median
of all models, without filtering by performance, and plot the results:

```{r, fig.width=6, fig.height=4}
lacerta_rep_ens_pred <- predict_raster(lacerta_rep_ens, climate_present_lac,
  metric_thresh = c("boyce_cont", 0.5),
  fun = c("mean", "median")
)

ggplot() +
  geom_spatraster(data = lacerta_rep_ens_pred, aes(fill = median)) +
  scale_fill_terrain_c()
```

Using DALEX

```{r, error = TRUE}
# create an explainer
explainer_lacerta_rep <- explain_tidysdm(lacerta_rep_ens)
```

```{r, error = TRUE}
# create feature importance plots
vip_ensemble <- model_parts(explainer = explainer_lacerta_rep)
```

```{r, error = TRUE}
plot(vip_ensemble)
```

```{r, error = TRUE}
# create partial dependence plots
pdp_bio05 <- model_profile(explainer_lacerta_rep, N = 500, variables = "bio05")
```

```{r}
plot(pdp_bio05)
```

Individual models:

```{r}
explainer_list <- explain_tidysdm(lacerta_rep_ens, by_workflow = TRUE)
```

```{r}
profile_list <- lapply(explainer_list, model_profile,
  N = 500,
  variables = "bio05"
)
```

```{r}
plot(profile_list)
```
DALEX currently does not work for `by_workflow = TRUE` with `repeat_ensemble`. We might think about if we would like this functionality in the future and how to do it. Perhaps an option to explore individual repeats, plotting average curve per repeat, or plotting average curve per model type across repeats?



# Repeat stack

Initial split

```{r}
lacerta_thin <- readRDS(system.file("extdata/lacerta_thin_all_vars.rds",
  package = "tidysdm"
))
```

```{r }
set.seed(100)
lacerta_initial <- spatial_initial_split(lacerta_thin,
  prop = 1 / 5, spatial_block_cv
)
autoplot(lacerta_initial)
```

```{r}
check_splits_balance(lacerta_initial, class)
```

```{r}
# select variables

lacerta_training <- training(lacerta_initial)

suggested_vars <- c("bio06", "bio05", "bio13", "bio14", "bio15")
# in lacerta_training, keep only columns named in suggested_vars and class
lacerta_training <- lacerta_training %>%
  dplyr::select(class, dplyr::all_of(suggested_vars))
```

```{r}
# empty object to store the simple ensembles that we will create
ensemble_list <- list()
set.seed(1234) # make sure you set the seed OUTSIDE the loop
for (i_repeat in 1:3) {
  # create folds
  lacerta_training_cv <- spatial_block_cv(lacerta_training, v = 5)
  # create a recipe
  lacerta_training_rec <- recipe(lacerta_training, formula = class ~ .)
  # create a workflow_set
  lacerta_training_models <-
    # create the workflow_set
    workflow_set(
      preproc = list(default = lacerta_training_rec),
      models = list(
        # the standard glm specs
        glm = sdm_spec_glm(),
        # maxent specs with tuning
        maxent = sdm_spec_maxent()
      ),
      # make all combinations of preproc and models,
      cross = TRUE
    ) %>%
    # tweak controls to store information needed later to create the ensemble
    option_add(control = control_ensemble_grid())

  # train the model
  lacerta_training_models <-
    lacerta_training_models %>%
    workflow_map("tune_grid",
      resamples = lacerta_training_cv, grid = 3,
      metrics = sdm_metric_set(), verbose = TRUE
    )
  
  # model stacking
  lacerta_stack <-
    stacks() %>%
    add_candidates(lacerta_training_models) %>%
    blend_predictions(
      metric = sdm_metric_set()
    ) %>% 
    fit_members()
  
  # make an ensemble and add it to the list
  ensemble_list[[i_repeat]] <- lacerta_stack
}
```

Now we can create a `repeat_stack` from the list:

```{r}
lacerta_rep_stack <- repeat_stack() %>% add_repeat(ensemble_list)
```

We can summarise the goodness of fit of models for each repeat with
`collect_metrics()`, but there is no `autoplot()` function for
`repeated_ensemble` objects.

We can then predict in the usual way. We will take the mean and median
of all models, without filtering by performance, and plot the results:

```{r, fig.width=6, fig.height=4}
lacerta_rep_stack_pred <- predict_raster(lacerta_rep_stack, climate_present_lac,
  fun = c("mean", "median")
)

ggplot() +
  geom_spatraster(data = lacerta_rep_stack_pred, aes(fill = median)) +
  scale_fill_terrain_c() +
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```

Plot individual stacks from the repeated stack

```{r}
prediction_st01 <- predict_raster(lacerta_rep_stack$.stack[[1]], climate_present,
                                     type = "prob"
)
ggplot() +
  geom_spatraster(data = prediction_st01, aes(fill = .pred_presence)) +
  scale_fill_terrain_c() +
  # plot presences used in the model
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))


prediction_st02 <- predict_raster(lacerta_rep_stack$.stack[[2]], climate_present,
                                     type = "prob"
)
ggplot() +
  geom_spatraster(data = prediction_st02, aes(fill = .pred_presence)) +
  scale_fill_terrain_c() +
  # plot presences used in the model
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))



prediction_st03 <- predict_raster(lacerta_rep_stack$.stack[[3]], climate_present,
                                     type = "prob"
)
ggplot() +
  geom_spatraster(data = prediction_st03, aes(fill = .pred_presence)) +
  scale_fill_terrain_c() +
  # plot presences used in the model
  geom_sf(data = lacerta_thin %>% filter(class == "presence"))
```

Then it would be great to be able to apply DALEX to explore the repeated
stacked ensemble, creating an explainer, feature importance plots,
response curves... But it does not work at the moment. 

```{r, error = TRUE}
# create an explainer
explainer_lacerta_rep_stack <- explain_tidysdm(lacerta_rep_stack)
```






