---
title: "Projecting your map"
author: "Andrea"
date: "2024-11-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Map projections in R

We start with a raster object that has a geographic coordinate reference system (CRS) and we want to project it to a different CRS. We will use the `terra` package to do this. As an example, we will use a map of the Iberian peninsula:


```{r}
library(pastclim)
download_dataset(dataset = "WorldClim_2.1_10m")
land_mask <-
  get_land_mask(time_ce = 1985, dataset = "WorldClim_2.1_10m")

# Iberia peninsula extension
iberia_poly <-
  terra::vect(
    "POLYGON((-9.8 43.3,-7.8 44.1,-2.0 43.7,3.6 42.5,3.8 41.5,1.3 40.8,0.3 39.5,
     0.9 38.6,-0.4 37.5,-1.6 36.7,-2.3 36.3,-4.1 36.4,-4.5 36.4,-5.0 36.1,
    -5.6 36.0,-6.3 36.0,-7.1 36.9,-9.5 36.6,-9.4 38.0,-10.6 38.9,-9.5 40.8,
    -9.8 43.3))"
  )

crs(iberia_poly) <- "+proj=longlat"

# crop the extent
land_mask <- crop(land_mask, iberia_poly)
land_mask <- mask(land_mask, iberia_poly)
```

We plot it with `tidyterra`:

```{r}
library(tidyterra)
library(ggplot2)
ggplot() +
  geom_spatraster(data = land_mask, aes(fill = land_mask_1985))
```
Now we project it. To define our projection we use a proj4 string, which provides information
on the type of projection, its parameters and the units of distance in which the new coordinates
will be expressed. we can use the website `projectionwizard.org` (https://link.springer.com/chapter/10.1007/978-3-319-51835-0_9) to find an appropriate equal
area projection for any region, and obtain its associated proj4 string.
In this case, we will use a Alberts Equal Area Conic projection centered on the Iberian peninsula. The proj4 string for this projection is:
```{r}
iberia_proj4 <- "+proj=aea +lon_0=-4.0429687 +lat_1=36.7790694 +lat_2=42.6258819 +lat_0=39.7024757 +datum=WGS84 +units=km +no_defs"
```
Note that we set the units to km so that have smaller numbers in the new axes.

For rasters (maps), we use the `terra` function `project` to change the CRS. We pass the raster object and the proj4 string as arguments:
```{r}
land_mask_proj <- terra::project(land_mask, y = iberia_proj4)
```

And replot it:
```{r}
ggplot() +
  geom_spatraster(data = land_mask_proj, aes(fill = land_mask_1985))
```
Note that the coordinate system has now changed. We can get the true coordinates using the terra::plot function:
```{r}
terra::plot(land_mask_proj)

```

We now need to bring in some points. they come in as lat long, so we use the appropriate proj4 string
for raw coordinates.
```{r}
library(tidysdm)
library(sf)
data(lacerta)
lacerta <- st_as_sf(lacerta, coords = c("longitude", "latitude"))
st_crs(lacerta) <- "+proj=longlat"
  #4326
```

Let's inspect the object:
```{r}
lacerta
```
We can see in the `geometry` column that are coordinates are in arc-degrees long and lat.

Now we project them to the same CRS as the raster, using the appropriate `sf` function
to project points:
```{r}
lacerta_proj <- st_transform(lacerta, iberia_proj4)
```

```{r}
lacerta_proj
```
Now the coordinates are in kilometers on the new axes.

Plot the projected points on top of the project map:
```{r}
ggplot() +
  geom_spatraster(data = land_mask_proj, aes(fill = land_mask_1985)) +
  geom_sf(data = lacerta_proj) + guides(fill="none")
```

The next step is to project the environemntal variables (layers of a raster). First we
extract the unprojected layers from `pastclim`
```{r}
download_dataset("WorldClim_2.1_10m")
climate_vars <- get_vars_for_dataset("WorldClim_2.1_10m")
climate_present <- pastclim::region_slice(
  time_ce = 1985,
  bio_variables = climate_vars,
  data = "WorldClim_2.1_10m",
  crop = iberia_poly
)
```
And now we project them
```{r}
climate_present_proj <- terra::project(climate_present, y = iberia_proj4)
```

Let's plot a layer with the points on top:
```{r}
ggplot() +
  geom_spatraster(data = climate_present_proj, aes(fill = bio01)) +
  geom_sf(data = lacerta_proj) + guides(fill="none") +
  scale_fill_gradientn(colors = terrain.colors(7), na.value = "transparent") 
```


